# Solutions for Worksheet 2 Exercises

2
```{r}
census <- read.csv(file="./data/1840-census-data.csv")[ , c("QualifyingAreaName", "Newspapers", "Newspapers_Daily", "Newspapers_Weekly", "Newspapers_SemiTriWeekly", "Periodicals", "PrintingOffices")]
census <- rename(census, county = QualifyingAreaName)
census_long <- gather(census, "publication_type", "count", 2:7)
census_long <- separate(census_long, county, into = c("county", "state"), sep = "\\, ")
census_long <- na.omit(census_long)
```

3
```{r}
census <- read_csv(file="./data/1840-census-data.csv")
census <- rename(census, county = QualifyingAreaName)
census <- separate(census, county, into = c("county", "state"), sep = "\\, ")
census <- group_by(census, state)
literate <- summarize(census, totalLit = sum(LiterateWhiteAge20andOver))
literate <- arrange(literate, desc(totalLit))

census <- read_csv(file="./data/1840-census-data.csv")
census <- rename(census, county = QualifyingAreaName)
census <- separate(census, county, into = c("county", "state"), sep = "\\, ")
census <- group_by(census, state)
literate <- summarize(census, totalLit = sum(LiterateWhiteAge20andOver) / sum(TotalPopulation))
literate <- arrange(literate, desc(totalLit))
```


# Solutions for Worksheet 3 Questions

What might we do if we wanted to see all of the rows that caused errors during the `separate` command? 

```{r}
View(filter(census, is.na(census$state)))
census %>% filter(is.na(.$state)) %>% View()
```

# Solutions for Worksheet 4 Questions

3. `([^"]*)` gets rid of everything before the quotation marks; `[ -]("[A-Za-z].*?")[ -]` finds everything within quotation marks. Once you get rid of the line breaks, `[^\"]*(?:\"([^\"]*)\")? ` does the trick.
4. `\b\w*([a-zA-z])\1{2,}\w*\b` or `\b\[A-Za-z]*([a-zA-z])\1{2,}\[A-Za-z]*\b` finds all the words with three or more repeated letters. 
5. `([A-Za-z]+)?([^Cc])(e)(i)(\S*)` finds the words with something ei; replace with `\1\2\4\3\5`

# Solutions for Worksheet 5

The question about creating new columns

```{r}
papers <- papers %>% 
  mutate(startDecade = str_sub(start, 1, 3)) %>%
  mutate(startDecade = as.numeric(paste(startDecade, 0, sep=""))) %>%
  mutate(endDecade = str_sub(end, 1,3)) %>%
  mutate(endDecade = as.numeric(paste(endDecade, 0, sep=""))) %>%
  unique()
```

Creating a percentage column in `topWords`/`titleWords`

```{r}

newPapers <- papers %>% 
  group_by(startDecade) %>%
  summarise(newPapers = n())

titleWords <- titleWords %>%
  left_join(newPapers, by = "startDecade") %>%
  mutate(percentage = count/newPapers) %>%
  arrange(startDecade, desc(percentage))


topWords <- function(x, decade1, decade2, nwords) {
  y <- x %>% 
    filter(startDecade >= decade1 & startDecade <= decade2) %>%
    group_by(startDecade) %>%
    summarise(newPapers = n())
  stopWords <- as_data_frame(c("the", "an", "and", "der", "die", 
                               "das", "und", "of", "in","aus","dem","or")) %>%
    rename(word = value)
  x %>%
    filter(startDecade >= decade1 & startDecade <= decade2) %>%
    unnest_tokens(word, title) %>%
    anti_join(stopWords) %>%
    group_by(startDecade, word) %>% 
    summarize(count = n()) %>%
    left_join(y, by = "startDecade") %>%
    mutate(percentage = count/newPapers) %>%
    arrange(startDecade, desc(percentage), desc(count)) %>%
    top_n(nwords)
}

```

Plotting top title words by year

```{r}

topWords <- function(x, year1, year2, nwords) {
  y <- x %>% 
    filter(start >= year1 & start <= year2) %>%
    group_by(start) %>%
    summarise(newPapers = n())
  stopWords <- as_data_frame(c("the", "an", "and", "der", "die", 
                               "das", "und", "of", "in","aus","dem","or")) %>%
    rename(word = value)
  x %>%
    filter(start >= year1 & start <= year2) %>%
    unnest_tokens(word, title) %>%
    anti_join(stopWords) %>%
    group_by(start, word, startDecade, end) %>% 
    summarize(count = n()) %>%
    ungroup() %>%
    group_by(start) %>% 
    left_join(y, by = "start") %>%
    mutate(percentage = count/newPapers) %>%
    arrange(start, desc(percentage), desc(count)) %>%
    top_n(nwords)
}

titleWords <- topWords(papers,1800,1950,3)

plot <- titleWords %>%
  top_n(3) %>% 
  ggplot(aes(x=start, y=percentage, color = word)) +
  geom_line() +
  geom_point(size = .3) +
  ggtitle(paste("Most Used Words in New Newspaper Titles by Decade, ", min(titleWords$start), "-", max(titleWords$start))) +
  labs(x="Decades",y="Percentage of Titles",fill="Word",caption="The top words used in the titles of new newspapers during the nineteenth century by decade") + 
  theme(plot.title = element_text(family = "Trebuchet MS", color="#666666", face="bold", size=18, hjust=0.5)) +
  theme(axis.title = element_text(family = "Trebuchet MS", color="#666666", face="bold", size=14)) + 
  theme(legend.title = element_text(family = "Trebuchet MS", color="#666666", face="bold", size=14)) +
  theme(legend.background = element_rect(color = "#efefef")) +
  theme(plot.caption = element_text(family = "Trebuchet MS", color="#666666", size=10, hjust = 0.5, margin = margin(15, 0, 15, 0))) +
  theme(axis.text = element_text(family = "Trebuchet MS", color="#aaaaaa", face="bold", size=10)) +
  theme(panel.background = element_rect(fill = "white")) +
  theme(panel.grid.major = element_line(color = "#efefef")) +
  theme(axis.ticks = element_line(color = "#efefef"))

ggplotly(plot)

```
