# Solutions for Worksheet 2 Exercises

2
```{r}
census <- read.csv(file="./data/1840-census-data.csv")[ , c("QualifyingAreaName", "Newspapers", "Newspapers_Daily", "Newspapers_Weekly", "Newspapers_SemiTriWeekly", "Periodicals", "PrintingOffices")]
census <- rename(census, county = QualifyingAreaName)
census_long <- gather(census, "publication_type", "count", 2:7)
census_long <- separate(census_long, county, into = c("county", "state"), sep = "\\, ")
census_long <- na.omit(census_long)
```

3
```{r}
census <- read_csv(file="./data/1840-census-data.csv")
census <- rename(census, county = QualifyingAreaName)
census <- separate(census, county, into = c("county", "state"), sep = "\\, ")
census <- group_by(census, state)
literate <- summarize(census, totalLit = sum(LiterateWhiteAge20andOver))
literate <- arrange(literate, desc(totalLit))

census <- read_csv(file="./data/1840-census-data.csv")
census <- rename(census, county = QualifyingAreaName)
census <- separate(census, county, into = c("county", "state"), sep = "\\, ")
census <- group_by(census, state)
literate <- summarize(census, totalLit = sum(LiterateWhiteAge20andOver) / sum(TotalPopulation))
literate <- arrange(literate, desc(totalLit))
```


# Solutions for Worksheet 3 Questions

What might we do if we wanted to see all of the rows that caused errors during the `separate` command? 

```{r}
View(filter(census, is.na(census$state)))
census %>% filter(is.na(.$state)) %>% View()
```

# Solutions for Worksheet 4 Questions

3. `([^"]*)` gets rid of everything before the quotation marks; `[ -]("[A-Za-z].*?")[ -]` finds everything within quotation marks. Once you get rid of the line breaks, `[^\"]*(?:\"([^\"]*)\")? ` does the trick.
4. `\b\w*([a-zA-z])\1{2,}\w*\b` or `\b\[A-Za-z]*([a-zA-z])\1{2,}\[A-Za-z]*\b` finds all the words with three or more repeated letters. 
5. `([A-Za-z]+)?([^Cc])(e)(i)(\S*)` finds the words with something ei; replace with `\1\2\4\3\5`